---
title: "smapr"
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/img/"
)
```

[![Build Status](https://travis-ci.org/earthlab/smapr.svg?branch=master)](https://travis-ci.org/earthlab/smapr) 

An R package for acquisition and processing of [NASA (Soil Moisture Active-Passive) SMAP data](http://smap.jpl.nasa.gov/)

## Installation

Development version:

```{r, eval = FALSE}
devtools::install_github("earthlab/smapr")
```

### Finding data

Data are hosted on an FTP server by the National Snow and Ice Data Center.
The `find_smap` function searches for specific data products and returns a data frame of available data.

```{r}
library("smapr")

files <- find_smap(id = "SPL3SMP", date = "2015-05-01", version = 3)
files
```

### Downloading data

Once found, data can be downloaded with `download_smap`.

```{r}
downloads <- download_smap(files)
downloads
```


### Extracting data

The `extract_smap` function extracts gridded data products (e.g., global soil moisture) and returns Raster* objects with the proper spatial projections etc.

```{r}
r <- extract_smap(downloads, name = 'Soil_Moisture_Retrieval_Data/soil_moisture')
r
```

### Plotting the data

Plotting is best done with the `raster` or `rasterVis` packages.

```{r}
raster::plot(r, main = "Soil Moisture: May 01, 2015")
```


### Inspecting data

The SMAP data are provided in HDF5 format, and in any one file there are actually multiple data sets, including metadata.
The `list_smap` function allows users to inspect the contents of downloaded data.

```{r}
list_smap(downloads, all = TRUE)
```

### Saving as a GeoTIFF

Users can save the rasters as GeoTIFFs with the `raster` package.

```{r, eval = FALSE}
raster::writeRaster(r, filename = "smap.tif")
```

### Running in Docker

To avoid dependency heck, we have made a Docker image available with smapr and all dependencies.

```{r, eval = FALSE, engine='bash'}
docker run -it mbjoseph/smapr bash
```
