---
title: "smapr"
output: github_document
---
[![Build Status](https://travis-ci.org/earthlab/smapr.svg?branch=master)](https://travis-ci.org/earthlab/smapr) 
[![codecov](https://codecov.io/gh/earthlab/smapr/branch/master/graph/badge.svg)](https://codecov.io/gh/earthlab/smapr)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/smapr)](https://cran.r-project.org/package=smapr)
[![Licence](https://img.shields.io/badge/licence-GPL--2-green.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.html)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](/commits/master)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/img/"
)
```


An R package for acquisition and processing of [NASA (Soil Moisture Active-Passive) SMAP data](http://smap.jpl.nasa.gov/)

## Installation

### Dependencies

To install smapr, you'll need the following R packages: 

- httr
- rappdirs
- raster
- rgdal
- rhdf5
- rvest
- utils
- xml2

Note that rhdf5 is a [Bioconductor](http://bioconductor.org/) package, and will not be installed by default when trying to install smapr via `install.packages`. 
For rhdf5 installation instructions see http://bioconductor.org/packages/release/bioc/html/rhdf5.html

To install from CRAN: 

```{r, eval = FALSE}
install.packages("smapr")
```


Development version:

```{r, eval = FALSE}
devtools::install_github("earthlab/smapr")
```


## Authentication

Access to the NASA SMAP data requires authentication through NASA's Earthdata portal. 
If you do not already have a username and password through Earthdata, you can register for an account here: https://urs.earthdata.nasa.gov/
You cannot use this package without an Earthdata account. 

Once you have an account, you need to pass your Earthdata username (`ed_un`) and password (`ed_pw`) as environmental variables that can be read from within your R session. 
There are a couple of ways to do this: 

- Use `Sys.setenv()` interactively in your R session to set your username and password (not including the `<` and `>`):

```{r, eval = FALSE}
Sys.setenv(ed_un = "<your username>", ed_pw = "<your password>")
```

- Use `Sys.setenv()` in your `.Rprofile` to set those environmental variables every time you load R. 

- Create a text file `.Renviron` in your home directory, which contains your username and password. 
If you don't know what your home directory is, execute `normalizePath("~/")` in the R console and it will be printed.
Be sure to include a new line at the end of the file or R will fail silently when loading it.

Example `.Renviron file` (note the new line at the end!):

```
ed_un=slkdjfsldkjfs
ed_pw=dlfkjDD124^

```

Once this file is created, restart your R session and you should now be able to access these environment variables (e.g., via `Sys.getenv(ed_un)`).



# SMAP data products

Multiple SMAP data products are provided by the NSIDC, and these products vary in the amount of processing. 
Currently, smapr supports the more processed level 3 and level 4 data products, which represent global daily composite and global three hourly modeled data products, respectively. 
NSIDC provides documentation for all SMAP data products on their [website](https://nsidc.org/data/smap/smap-data.html), and we provide a summary of level 3 and level 4 products below. 

| Dataset id | Description                                         | Resolution |
|------------|-----------------------------------------------------|------------|
| SPL3FTA    | Radar Northern Hemisphere Daily Freeze/Thaw State   | 3 km       |
| SPL3SMA    | Radar Global Daily Soil Moisture                    | 3 km       |
| SPL3SMP    | Radiometer Global Soil Moisture                     | 36 km      |
| SPL3SMAP   | Radar/Radiometer Global Soil Moisture               | 9 km       |
| SPL4SMAU   | Surface/Rootzone Soil Moisture Analysis Update      | 9 km       | 
| SPL4SMGP   | Surface/Rootzone Soil Moisture Geophysical Data     | 9 km       |
| SPL4SMLM   | Surface/Rootzone Soil Moisture Land Model Constants | 9 km       |
| SPL4CMDL   | Carbon Net Ecosystem Exchange                       | 9 km       |

### Finding SMAP data

Data are hosted on a server by the National Snow and Ice Data Center. 
The find_smap function searches for specific data products and returns a data frame of available data.
As data mature and pass checks, versions advance. 
At any specific time, not all versions of all datasets for all dates may exist. 
For the most up to date overview of dataset versions, see the NSIDC SMAP data version [webpage](https://nsidc.org/data/smap/smap-data.html).

```{r}
library(smapr)
library(raster)
available_data <- find_smap(id = "SPL3SMAP", date = "2015-05-25", version = 3)
str(available_data)
```

### Downloading and inspecting SMAP data

Given a data frame produced by `find_smap`, `download_smap` downloads the data onto the local file system. 
Unless a directory is specified as an argument, the data are stored in the user's cache. 

```{r}
downloads <- download_smap(available_data)
str(downloads)
```

The SMAP data are provided in HDF5 format, and in any one file there are actually multiple data sets, including metadata. 
The `list_smap` function allows users to inspect the contents of downloaded data at a high level (`all = FALSE`) or in depth (`all = TRUE`). 

```{r}
list_smap(downloads, all = FALSE)
list_smap(downloads, all = TRUE)
```

### Extracting gridded data products

The `extract_smap` function extracts gridded data products (e.g., global soil moisture) and returns Raster* objects with the proper spatial projections etc. 
If more than one file has been downloaded and passed into the first argument, `extract_smap` extracts all of the rasters and returns a RasterStack.

```{r, fig.align='center', fig.width=8, fig.height=7}
sm_raster <- extract_smap(downloads, "Soil_Moisture_Retrieval_Data/soil_moisture")
plot(sm_raster, main = "Level 3 soil moisture: May 25, 2015")
```

## Saving GeoTIFF output

The raster stack can be saved as a GeoTIFF using the `writeRaster` function from the raster pacakge. 

```{r}
writeRaster(sm_raster, "wgs84_ft.tif")
```

```{r, echo = FALSE, results='hide'}
# cleanup
file.remove("wgs84_ft.tif")
```

### Running in Docker

To avoid dependency heck, we have made a Docker image available with smapr and all dependencies.

```{r, eval = FALSE, engine='bash'}
docker run -it earthlab/smapr bash
```

Bug reports, issues, and feature requests can be submitted via GitHub: https://github.com/earthlab/smapr
