---
title: "smapr"
output: github_document
---
[![Build Status](https://travis-ci.org/earthlab/smapr.svg?branch=master)](https://travis-ci.org/earthlab/smapr) 
[![codecov](https://codecov.io/gh/earthlab/smapr/branch/master/graph/badge.svg)](https://codecov.io/gh/earthlab/smapr)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/smapr)](https://cran.r-project.org/package=smapr)
[![Licence](https://img.shields.io/badge/licence-GPL--2-blue.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.html)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](/commits/master)


 

 

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/img/"
)
```




An R package for acquisition and processing of [NASA (Soil Moisture Active-Passive) SMAP data](http://smap.jpl.nasa.gov/)

## Installation

### Dependencies

To install smapr, you'll need the following R packages: 

- curl
- httr
- rappdirs
- raster
- rgdal
- rhdf5
- utils

Note that rhdf5 is a [Bioconductor](http://bioconductor.org/) package, and will not be installed by default when trying to install smapr via `install.packages`. 
For rhdf5 installation instructions see http://bioconductor.org/packages/release/bioc/html/rhdf5.html

To install from CRAN: 

```{r, eval = FALSE}
install.packages("smapr")
```


Development version:

```{r, eval = FALSE}
devtools::install_github("earthlab/smapr")
```


## Authentication

Access to the NASA SMAP data requires authentication through NASA's Earthdata portal. 
If you do not already have a username and password through Earthdata, you can register for an account here: https://urs.earthdata.nasa.gov/
You cannot use this package without an Earthdata account. 

Once you have an account, you need to pass your Earthdata username (`ed_un`) and password (`ed_pw`) as environmental variables that can be read from within your R session. 
There are a couple of ways to do this: 

A) Use `Sys.setenv()` interactively in your R session to set your username and password (not including the `<` and `>`):

```{r, eval = FALSE}
Sys.setenv(ed_un = "<your username>", ed_pw = "<your password>")
```

B) Use `Sys.setenv()` in your `.Rprofile` to set those environmental variables every time you load R. 

### Finding data

Data are hosted on an FTP server by the National Snow and Ice Data Center.
The `find_smap` function searches for specific data products and returns a data frame of available data.

```{r}
library("smapr")

files <- find_smap(id = "SPL3SMP", date = "2015-05-01", version = 4)
files
```

### Downloading data

Once found, data can be downloaded with `download_smap`.

```{r}
downloads <- download_smap(files)
downloads
```


### Extracting data

The `extract_smap` function extracts gridded data products (e.g., global soil moisture) and returns Raster* objects with the proper spatial projections etc.

```{r}
r <- extract_smap(downloads, name = 'Soil_Moisture_Retrieval_Data_AM/soil_moisture')
r
```

### Plotting the data

Plotting is best done with the `raster` or `rasterVis` packages.

```{r}
raster::plot(r, main = "AM Soil Moisture: May 01, 2015")
```


### Inspecting data

The SMAP data are provided in HDF5 format, and in any one file there are actually multiple data sets, including metadata.
The `list_smap` function allows users to inspect the contents of downloaded data.

```{r}
list_smap(downloads, all = TRUE)
```

### Saving as a GeoTIFF

Users can save the rasters as GeoTIFFs with the `raster` package.

```{r, eval = FALSE}
raster::writeRaster(r, filename = "smap.tif")
```

### Running in Docker

To avoid dependency heck, we have made a Docker image available with smapr and all dependencies.

```{r, eval = FALSE, engine='bash'}
docker run -it earthlab/smapr bash
```
