---
title: "Introduction to the smapr package"
author: "Maxwell B. Joseph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the smapr package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(purl = NOT_CRAN)
```

```{r load-smapr}
library(smapr)
library(raster)
```


This vignette outlines a basic use scenario for smapr.
We will acquire and process 
[NASA (Soil Moisture Active-Passive) SMAP data](http://smap.jpl.nasa.gov/), 
and generate some simple visualizations. 
We'll use the global level 4 [SPL4SMAU](https://nsidc.org/data/SPL4SMAU) 
(Surface/Rootzone Soil Moisture Analysis Update) data product. 

## Finding data

To find out which SMAP data are available, we'll use the `find_smap` function, 
which takes a data set ID, date(s) to search, and a dataset version. 
The SPL4SMAU data product is on version 3 (see https://nsidc.org/data/SPL4SMAU), 
and we'll search for data from June 01, 2018.

```{r find-data, eval = NOT_CRAN}
available_data <- find_smap(id = 'SPL4SMAU', dates = '2018-06-01', version = 3)
```

This returns a data frame, where every row is one data file that is available 
on NASA's servers. 

## Downloading data

To download the data, we can use `download_smap`:

```{r download-data, eval = NOT_CRAN}
local_files <- download_smap(available_data)
```

Now we have 8 data files, which correspond to different times of day on our 
search date, as indicated by the file names: 

```{r print-filenames, eval = NOT_CRAN}
local_files$name[1:2]
```

## Exploring data

Each file that we downloaded is an HDF5 file with multiple datasets bundled 
together. 
To list all of the data in a file we can use `list_smap`. 
By default, if we give `list_smap` a data frame of local files, it will 
return a list of data frames. 
Because all of these data files are of the same data product, using `list_smap`
on one file (e.g., the first) will tell us what's available in all of the files: 

```{r list-smap, eval = NOT_CRAN}
list_smap(local_files[1, ])
```

To dig deeper, we can use the `all` argument to `list_smap`:

```{r list-more-smap, eval = NOT_CRAN}
list_smap(local_files[1, ], all = TRUE)
```

Looking at this output, we can conclude that the file contains multiple arrays 
(notice the `dim` column). 
These arrays correspond to things like estimated root zone soil moisture 
(`/Analysis_Data/sm_rootzone_analysis`), estimated surface soil moisture
(`/Analysis_Data/sm_surface_analysis`), and estimated surface temperature 
(`/Analysis_Data/surface_temp_analysis`). 
See https://nsidc.org/data/smap/spl4sm/data-fields#sm_surface_analysis for more 
detailed information on what these datasets represent and how they were 
generated.

## Extracting data

The datasets that we are interested in are spatial grids. 
The `smapr` package can extract these data into `raster` objects with the 
`extract_smap` function, which takes a dataset name as an argument.
These names are paths that can be generated from the output of `list_smap`. 
For example, if we want to get rootzone soil moisture, we can see a dataset 
with name `sm_rootzone_analysis` in group `/Analysis_Data`, so that the path 
to the dataset is `/Analysis_Data/sm_rootzone_analysis`: 

```{r extract-data, eval = NOT_CRAN}
sm_raster <- extract_smap(local_files, '/Analysis_Data/sm_rootzone_analysis')
```

This will extract all of the data in the data frame `local_files`, generating 
a RasterBrick with one layer per file:

```{r print-raster, eval = NOT_CRAN}
sm_raster
```

We can visualize root zone soil moisture at different times using the raster 
package:

```{r plot-raster, fig.width=12, fig.height=8, eval = NOT_CRAN}
plot(sm_raster)
```
